
---
title: 'Monthly Readers Interactions'
output: rmarkdown::github_document
---

TODOs

* Move process to SWAP noteboook.
* Create intermediate hive table to help automate and streamline the process.
* Note monthly values are based on calendar month. They are no longer normalized due to 30 days. 
* Note: Various IE adjustments and corrections made for previous years. (IEPKIRAF: since May 20, 2018, IE7PKIRAF: January 1-May 19, 2018, IE7PKIRAF & iOS mainpage: ...-Dec 31, 2017)


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, tidy=TRUE,
fig.path = "figures/md_figs/")
```

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(scales)
library(reshape2)
```


## Monthly Pageviews

```{r eval=FALSE}
#Get overall pageviews in the month (all wikis, global) with IE correction applied for post May 20, 2018 dates. Other corrections applied for earlier dates. 
query <- "
SELECT year, month, day, CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
SUM(IF(access_method = 'mobile app', view_count, null)) AS Apps,
SUM(IF(access_method = 'desktop', view_count, null)) AS Desktop,
SUM(IF(access_method = 'mobile web', view_count, null)) AS MobileWeb,
SUM(view_count) as Total
FROM wmf.pageview_hourly
WHERE ((year = 2018 AND month >= 5) OR (year >= 2019))
AND agent_type='user'
AND NOT (country_code IN ('PK', 'IR', 'AF') -- https://phabricator.wikimedia.org/T157404#3194046
AND user_agent_map['browser_family'] = 'IE') -- https://phabricator.wikimedia.org/T193578#4300284
GROUP BY year, month, day ORDER BY year, month, day LIMIT 1000;
"
results <- collect(sql(query))
#add results to pageviews.tsv file which includes corrected data for pre-May 2018 dates.
#TODO: Add other queries with corrections for earlier dates to automate this process
```


```{r}
#Read in pageviews data with all corrected data (dating back to 2015-04-01)
pageviews_all <- read.delim("data/pageviews.tsv", sep = "\t")
pageviews_all$date <- as.Date(pageviews_all$date, format = "%Y-%m-%d")
```

Find total pageviews per month with yoy changes
```{r}
pageviews_all_monthly <- pageviews_all %>%
  mutate(date = floor_date(date, "month")) %>%
  group_by(date) %>%
  summarise(monthly_views = sum(as.numeric(total))) %>%
  arrange(date) %>%
  mutate(yearOverYear= monthly_views/lag(monthly_views,12) -1) %>%
  mutate(type = "pageviews")

tail(pageviews_all_monthly)
```

Create time series chart of monthly pageviews 
```{r}
p <- ggplot(pageviews_all_monthly, aes(x=date, y = monthly_views)) +  
  geom_line(color = 'blue', size = 1.5) +
  geom_vline(xintercept = as.numeric(as.Date("2018-04-01")),
             linetype = "dashed", color = "black") +
  geom_text(aes(x=as.Date('2018-04-01'), y=16E9, label="Page previews deployment completed"), size=3.5, vjust = -1.2, angle = 90, color = "black") +
  scale_y_continuous("pageviews per month (in billions)", labels = polloi::compress) +
  scale_x_date("Date", labels = date_format("%b %Y"), date_breaks = "2 months", limits= as.Date(c("2017-01-01", "2019-05-01"))) +
  labs(title = "Monthly Wikimedia pageviews") +
  ggthemes::theme_tufte(base_size = 12, base_family = "Gill Sans") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        panel.grid = element_line("gray70"))

ggsave(filename="Monthly pageviews monthly time series.png", plot = p, path = "figures", units = "in", dpi = 150, height = 6, width = 10, limitsize = FALSE)  
p
```


### Review pageviews by access method

Calculate desktop pageview and year over year comparison

```{r}
pageviews_desktop <- pageviews_all %>%
  mutate(date = floor_date(date, "month")) %>%
  select(c(1,3))  %>% #select desktop and date column
  group_by(date) %>%
  summarise(monthly_views = sum(as.numeric(desktop))) %>%
  arrange(date) %>%
  mutate(yearOveryear = monthly_views/lag(monthly_views,12) -1)

tail(pageviews_desktop)
```


Calculate mobile web pageview and year over year comparison
 
```{r}
pageviews_mobileweb <- pageviews_all  %>%
  mutate(date = floor_date(date, "month")) %>%
  select(c(1,4))  %>% #select mobile web and date column
  group_by(date) %>%
  summarise(monthly_views = sum(as.numeric(mobileweb))) %>%
  arrange(date) %>%
  mutate(yearOveryear = monthly_views/lag(monthly_views,12) -1)

tail(pageviews_mobileweb)
```
 

Quality check: Review all access methods together

```{r}
pageviews_byaccess <- pageviews_all %>%
  select(-c(5)) %>% #remove total column as it's not needed for this chart
  melt(id.vars = c("date"), measure.vars = c("apps", "desktop", "mobileweb"),
       variable.name = "access_method", value.name = 'views')

p <- ggplot(pageviews_byaccess, aes(x=date, y = (views/1000000), color = access_method)) +  
  geom_line() +
  geom_vline(xintercept = as.numeric(as.Date("2018-12-25")),
             linetype = "dashed", color = "blue") +
  geom_text(aes(x=as.Date('2018-12-25'), y=250, label="Christmas"), size=4, vjust = -1.2, angle = 90, color = "black") +
  scale_y_continuous("daily pageviews (in millions)") +
  scale_x_date("Date", labels = date_format("%Y-%m-%d"), date_breaks = "2 months", limits= as.Date(c("2018-01-01", "2019-05-01"))) +
  labs(title = "Total daily Wikimedia pageviews by access method") +
  ggthemes::theme_tufte(base_size = 12, base_family = "Gill Sans") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1),
        panel.grid = element_line("gray70"))


ggsave(filename="Daily pageviews byaccess method.png", plot = p, path = "figures", units = "in", dpi = 192, height = 6, width = 10, limitsize = FALSE)  
p
```


Create YOY Chart comparing the current fiscal year with the previous fical year.

```{r}
#Various adjustments to allow for fiscal year comparison.
pageviews_all_monthly_fy <- pageviews_all_monthly %>%
  filter(date >= '2017-07-01') %>%
  mutate(fiscal_year = ifelse(date >= '2017-07-01' & date < '2018-07-01', 'FY 2017/2018', 'FY 2018/2019'),
         MonthN =as.factor(format(as.Date(date),"%m")),
         Month = months(as.Date(date), abbreviate=TRUE))

pageviews_all_monthly_fy$MonthN = factor(pageviews_all_monthly_fy$MonthN, levels=c("07","08", "09", "10", "11", "12", "01", "02", "03", "04", "05", "06"))
  

p <- ggplot(pageviews_all_monthly_fy, aes(x=MonthN, y = monthly_views, group = fiscal_year, color = fiscal_year, linetype = fiscal_year)) +  
  geom_line(size = 1) +
  geom_vline(xintercept = 10,
             linetype = "dashed", color = "black") +
  geom_text(aes(x=10, y=15.5E9, label="Page previews deployment completed (April 2018)"), size=4, vjust = -1.2, angle = 90, color = "black") +
  scale_y_continuous("pageviews per month (in billions)", labels = polloi::compress) +
  scale_x_discrete(breaks = pageviews_all_monthly_fy$MonthN, labels = pageviews_all_monthly_fy$Month )+
  labs(title = "Wikimedia pageviews year-over-year comparison, July 2017-May 2019") +
  ggthemes::theme_tufte(base_size = 14, base_family = "Gill Sans") +
  theme(plot.title = element_text(hjust = 0.5),
        legend.title = element_blank(),
        legend.position = "bottom",
        panel.grid = element_line("gray70"),
        legend.key.width=unit(1.5,"cm")) +
  xlab("Month") +
  scale_linetype_manual(breaks=c('FY 2017/2018','FY 2018/2019'), values=c(2,1))

ggsave(filename="Pageviews_monthly_yoy.png", plot = p, path = 'figures', units = "in", dpi = 192, height = 6, width = 10, limitsize = FALSE)  
p
```


## Monthly Previews on Desktop Wikipedia

```{r eval=FALSE}
#Get daily page previews seen on desktop Wikipedia since deployment in April 2018
query <- "
SELECT year, month, day, CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
  SUM(view_count) AS previews_seen
  FROM wmf.virtualpageview_hourly 
  WHERE (year = 2018 AND month >=4) OR (year = 2019)
  GROUP BY year, month, day
  ORDER BY year, month, day LIMIT 10000;
"
results <- collect(sql(query))
save(results, file="Data/page_previews.tsv")
```

```{r}
previews <- read.delim("data/page_previews.tsv", sep = "\t")
previews$date <- as.Date(previews$date, format = "%Y-%m-%d")

previews_all_monthly <- previews %>%
  select(-c(1,2,3)) %>%
  mutate(date = floor_date(date, "month"))  %>%
  filter(date <= '2019-05-01') %>% #Filter out incomplete months
  group_by(date) %>%
  summarise(monthly_views = sum(as.numeric(previews_seen))) %>%
  arrange(date) %>%
  mutate(yearOverYear= monthly_views/lag(monthly_views,12) -1) %>%
  mutate(type = "previews")

tail(previews_all_monthly)
```

Chart daily page previews times series 

```{r}

previews_all_daily <- previews %>% 
  filter(date >= '2018-04-17')  #Filter out dates that occured before page previews rollout was completed.

p <- ggplot(previews_all_daily, aes(x=date, y = previews_seen)) +  
  geom_line(color = 'red', size = 1) +
  geom_vline(xintercept = as.numeric(as.Date("2018-04-17")),
             linetype = "dashed", color = "black")+
  geom_vline(xintercept = as.numeric(as.Date("2018-12-25")),
             linetype = "dashed", color = "black")+
  geom_text(aes(x=as.Date('2018-12-25'), y=50E6, label="Christmas"), size=4, vjust = -1.2, angle = 90, color = "black") +
  geom_text(aes(x=as.Date('2018-04-17'), y=55E6, label="Page previews deployment completed"), size=4, vjust = -1.2, angle = 90, color = "black") +
  scale_y_continuous("seen previews (shown for at least 1 sec)", labels = polloi::compress) +
  scale_x_date("Date", labels = date_format("%b %Y"), date_breaks = "1 month") +
  labs(title = "Daily seen previews on Wikipedia Desktop") +
  ggthemes::theme_tufte(base_size = 12, base_family = "Gill Sans") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1),
        plot.title = element_text(hjust = 0.5),
        panel.grid = element_line("gray70"))

ggsave(filename="Daily page previews seen.png", plot = p, path = "figures", units = "in", dpi = 192, height = 6, width = 10, limitsize = FALSE)  
p
```

# Monthly Interactions (Pageviews + Seen Previews)

Combine pageviews and previews data into table

```{r}
interactions <- rbind(pageviews_all_monthly, previews_all_monthly) 
```

Create stacked bar charts of page interactions

```{r}
p <- ggplot(interactions, aes(x=date, y = monthly_views, fill = forcats::fct_rev(type))) +  
  geom_col() +
  scale_y_continuous("seen previews (shown for at least 1 sec)\nand pageviews (in billions)",labels = polloi::compress) +
  scale_x_date("Date", labels = date_format("%Y-%m"), date_breaks = "3 months") +
  geom_vline(xintercept = as.numeric(as.Date("2018-04-01")),
             linetype = "dashed", color = "blue") +
  geom_text(aes(x=as.Date('2018-04-01'), y=8E9, label="Page Previews Deployment Completed"), size=4, vjust = -1.2, angle = 90, color = "black") +
  labs(title = "Pageviews and seen previews per calendar month") +
  ggthemes::theme_tufte(base_size = 14, base_family = "Gill Sans") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1),
        panel.grid = element_line("gray70"),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=14))

ggsave(filename="Pageviews and interactions_StackedBar.png", plot = p, path = 'figures', units = "in", dpi = 192, height = 6, width = 10, limitsize = FALSE)  
p
```

Calculate total monthly interactions (sum of pageviews and previews) on desktop

```{r}
interactions_total <- interactions %>%
  filter(date >= "2018-04-01") %>% #filter to first month previews rolled out.
  group_by(date) %>%
  summarise(interactions = sum(monthly_views)) %>%
  arrange(date) %>%
  mutate(yearOverYear= interactions/lag(interactions,12) -1) 

knitr::kable(interactions_total)
```



