
---
title: 'Readers Diversity Metrics'
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, tidy=TRUE, fig.path = "figures/README_figs/")
```

```{r}
library(tidyverse)
library(ggplot2)
library(lubridate)
library(scales)
library(reshape2)
```

# Interactions by Economic Region

## Global South pageviews

```{r eval=FALSE}
# Get pageviews by GS / GN / unknown for post May 2018 data. Other timespans need different corrections
# IE correction: https://phabricator.wikimedia.org/T157404#3194046, https://phabricator.wikimedia.org/T193578#4300284

query <- "
SELECT date,
countries.economic_region AS region,
SUM(pageviews_country) AS pageviews
FROM (
  SELECT year, month, day, 
  CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
  country_code,
  SUM(view_count) AS pageviews_country
  FROM wmf.pageview_hourly
  WHERE ((year = 2018 AND month >= 5) OR year >= 2019)
  AND agent_type='user'
  AND NOT (country_code IN ('PK', 'IR', 'AF') -- https://phabricator.wikimedia.org/T157404#3194046
  AND user_agent_map['browser_family'] = 'IE') -- https://phabricator.wikimedia.org/T193578#4300284
  GROUP BY year, month, day, country_code) AS bydatecountry
JOIN canonical_data.countries AS countries
ON bydatecountry.country_code = countries.iso_code
GROUP BY date, countries.economic_region
ORDER BY date, region LIMIT 10000;
"
results <- collect(sql(query))
save(results, file="Data/global_south_pageviews.tsv")

```

```{r eval=FALSE}
# Get pageviews by GS / GN / unknown with IE7PKIRAF Correction for January 1-May 19, 2018 dates.  https://phabricator.wikimedia.org/T157404#3194046
query <- "
SELECT date,
countries.economic_region AS region,
SUM(pageviews_country) AS pageviews
FROM (
  SELECT year, month, day, 
  CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
  country_code,
  SUM(view_count) AS pageviews_country
  FROM wmf.pageview_hourly
  WHERE year = 2018 AND ((month < 5) OR (month = 5 AND day <=19))
  AND agent_type='user'
  AND NOT (country_code IN ('PK', 'IR', 'AF')
  AND user_agent_map['browser_family'] = 'IE' AND user_agent_map['browser_major'] = 7)
  GROUP BY year, month, day, country_code) AS bydatecountry
JOIN canonical_data.countries AS countries
ON bydatecountry.country_code = countries.iso_code
GROUP BY date, countries.economic_region
ORDER BY date, region LIMIT 10000;
"
results <- collect(sql(query))
save(results, file="Data/global_south_pageview_Jan-May2018.tsv")

#Add/Reconcile global_south_pageview_Jan-May2018.tsv to global_south_pageviews.tsv 
#TODO: 
#Streamline and add code to automate query and data combining process
#Collect data with necessary corrections for 2015-2017 for historical analysis
```


```{r}
pageviews_gs_gn <- read.delim("data/global_south_pageviews.tsv", sep = "\t", stringsAsFactors =FALSE)
pageviews_gs_gn$date <- as.Date(pageviews_gs_gn$date, format = "%Y-%m-%d")
```

Calculate monthly and YoY pageviews for the global south region

```{r}
gs_pageviews_monthly <- pageviews_gs_gn %>%
  mutate(date = floor_date(date, "month")) %>%
  filter(date <= '2019-04-01',
         region == 'Global South') %>% #filter out May due to incomplete data. Look only at Global South
  group_by(date) %>%
  summarise(monthly_views = sum(as.numeric(pageviews))/1E9) %>%
  arrange(date) %>%
  mutate(yearOverYear= monthly_views/lag(monthly_views,12) -1)  %>%
  mutate(type = "pageviews")

tail(gs_pageviews_monthly)
```


## Global South seen previews

```{r eval=FALSE}
# Get seen previews by GS / GN / unknown
query <- "
SELECT date,
countries.economic_region AS region,
SUM(previews_seen_country) AS previews_seen
FROM (
  SELECT year, month, day, 
  CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
  country_code,
  SUM(view_count) AS previews_seen_country
  FROM wmf.virtualpageview_hourly 
  WHERE (year = 2018 AND month >=4) OR (year = 2019)
  GROUP BY year, month, day, country_code) AS bydatecountry
JOIN canonical_data.countries AS countries
ON bydatecountry.country_code = countries.iso_code
GROUP BY date, countries.economic_region
ORDER BY date, region LIMIT 10000
"
results <- collect(sql(query))
save(results, file="Data/global_south_previews.tsv")
```


```{r}
previews_gs_gn <- read.delim("data/global_south_previews.tsv", sep = "\t", stringsAsFactors =FALSE)
previews_gs_gn$date <- as.Date(previews_gs_gn$date, format = "%Y-%m-%d")
```

Calculate monthly and YoY previews on desktop Wikipedia for the global south region

```{r}
gs_previews_monthly <- previews_gs_gn %>%
  mutate(date = floor_date(date, "month")) %>%
  filter(date <= '2019-04-01',
         region == 'Global South') %>% #filter out May due to incomplete data. Look only at Global South
  group_by(date) %>%
  summarise(monthly_views = sum(as.numeric(previews_seen))/1E9) %>%
  arrange(date) %>%
  mutate(yearOverYear= monthly_views/lag(monthly_views,12) -1)  %>%
  mutate(type = "previews")

tail(gs_previews_monthly)
```

## Monthly Interactions in Global South (Pageviews + Seen Previews)

Create chart of interactions brokend down by pageviews and previews

```{r}
gs_interactions <- rbind(gs_pageviews_monthly, gs_previews_monthly) 

#Stacked bar chart to compare

p <- ggplot(gs_interactions, aes(x=date, y = monthly_views, fill = forcats::fct_rev(type))) +  
  geom_col() +
  scale_y_continuous("seen previews (shown for at least 1 sec)\nand pageviews (in billions)", breaks=c(seq(0, 20, by = 2.5))) +
  scale_x_date("Date", labels = date_format("%Y-%m"), date_breaks = "3 months") +
  geom_vline(xintercept = as.numeric(as.Date("2018-04-01")),
             linetype = "dashed", color = "blue") +
  geom_text(aes(x=as.Date('2018-04-01'), y=2.5, label="Page Previews Deployment Completed"), size=4, vjust = -1.2, angle = 90, color = "black") +
  labs(title = "Global south pageviews and seen previews per calendar month") +
  ggthemes::theme_tufte(base_size = 14, base_family = "Gill Sans") +
  theme(axis.text.x=element_text(angle = 45, hjust = 1),
        panel.grid = element_line("gray70"),
        legend.position="bottom",
        legend.title=element_blank(),
        legend.text=element_text(size=14))

ggsave(filename="Global South pageviews and interactions_StackedBar.png", plot = p, path = 'figures', units = "in", dpi = 192, height = 6, width = 10, limitsize = FALSE)  
p
```

Calculate monthly interactions (sum of pageviews and previews) and YoY changes in global south

```{r}
gs_interactions_total <- gs_interactions %>%
  filter(date >= "2018-04-01") %>% #filter to first month previews rolled out.
  group_by(date) %>%
  summarise(interactions = sum(monthly_views)) %>%
  arrange(date) %>%
  mutate(yearOverYear= interactions/lag(interactions,12) -1) 

knitr::kable(gs_interactions_total)
```


# Mobile-Heavy Wikis Page Interactions

```{r eval=FALSE}
# Get pageviews for identified mobile heavy countires for post May 2018 data. Other timespans need different corrections
# IE correction: https://phabricator.wikimedia.org/T157404#3194046, https://phabricator.wikimedia.org/T193578#4300284
# Mobile heave wikis defined at: https://github.com/wikimedia-research/canonical-data/blob/master/mobile_heavy_wikis.csv

query <- "SELECT year, month, day, CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
 SUM(IF (FIND_IN_SET (project, 'ar.wikipedia, arz.wikipedia, az.wikipedia, bn.wikipedia, 
	fa.wikipedia, gu.wikipedia, hi.wikipedia, id.wikipedia, kn.wikipedia,
	ky.wikipedia, ml.wikipedia, mr.wikipedia, ms.wikipedia, pt.wiktionary,
	sq.wikipedia, sw.wikipedia, ta.wikipedia, th.wikipedia, tl.wikipedia, 
	zh.wikiquote') > 0, view_count, 0)) AS mh_views
FROM wmf.pageview_hourly
WHERE ((year = 2018 AND month >=5) OR (year = 2019))
AND agent_type='user' 
AND NOT (country_code IN ('PK', 'IR', 'AF')
AND user_agent_map['browser_family'] = 'IE' AND user_agent_map['browser_major'] = 7)
GROUP BY year, month, day ORDER BY year, month, day LIMIT 1000;
"
results <- collect(sql(query))
save(results, file="Data/mobile_heavy_pageviews.tsv")

#TODO: See if you can adjust query to use canonical_data.mobile_heavy_wikis
```


```{r eval=FALSE}
# Get pageviews for identified mobile heavy countires for January 1-May 19, 2018 dates. Other timespans need different corrections   https://phabricator.wikimedia.org/T157404#3194046
query <- "
SELECT year, month, day, CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
 SUM(IF (FIND_IN_SET (project, 'ar.wikipedia, arz.wikipedia, az.wikipedia, bn.wikipedia, 
	fa.wikipedia, gu.wikipedia, hi.wikipedia, id.wikipedia, kn.wikipedia,
	ky.wikipedia, ml.wikipedia, mr.wikipedia, ms.wikipedia, pt.wiktionary,
	sq.wikipedia, sw.wikipedia, ta.wikipedia, th.wikipedia, tl.wikipedia, 
	zh.wikiquote') > 0, view_count, 0)) AS mh_views
-- Filter for mobile heavy wikis https://github.com/wikimedia-research/canonical-data/blob/master/mobile_heavy_wikis.csv
FROM wmf.pageview_hourly
WHERE ((year = 2018 AND month >=5) OR (year = 2019))
AND agent_type='user' 
AND NOT (country_code IN ('PK', 'IR', 'AF') -- https://phabricator.wikimedia.org/T157404#3194046
AND user_agent_map['browser_family'] = 'IE') -- https://phabricator.wikimedia.org/T193578#4300284
GROUP BY year, month, day ORDER BY year, month, day LIMIT 1000;
"
results <- collect(sql(query))
save(results, file="Data/mobile_heavy_pageviews_Jan-May2018.tsv")

#Add/Reconcile mobile_heavy_pageviews_Jan-May2018.tsv to mobile_heavy_pageviews.tsv 
#TODO: 
#Streamline and add code to automate query and data combining process
#Collect data with necessary corrections for 2015-2017 for historical analysis
```


## Mobile-Heavy Pageviews

```{r}
mobile_heavy_pageviews <- read.delim("data/mobile_heavy_pageviews.tsv", sep = "\t", stringsAsFactors =FALSE)
mobile_heavy_pageviews$date <- as.Date(mobile_heavy_pageviews$date, format = "%Y-%m-%d")
```

Calculate monthly pageviews and YoY changes on mobile-heavy wikis
```{r}
mh_pageviews_monthly <- mobile_heavy_pageviews %>%
  mutate(date = floor_date(date, "month")) %>%
  filter(date <= '2019-04-01') %>% #filter out months with incomplete date
  select(-c(1,2,3)) %>%
  group_by(date) %>% #filter out May due to incomplete data. Look only at Global South
  summarise(monthly_views = sum(as.numeric(mh_views))/1E6) %>%
  arrange(date) %>%
  mutate(yearOverYear= monthly_views/lag(monthly_views,12) -1)  %>%
  mutate(type = "pageviews")

tail(mh_pageviews_monthly)
```
  
## Mobile-Heavy Previews

```{r eval=FALSE}
# Get seen previews by GS / GN / unknown
query <- "
SELECT year, month, day, CONCAT(year,'-',LPAD(month,2,'0'),'-',LPAD(day,2,'0')) AS date,
  SUM(view_count) AS all,
SUM(IF (FIND_IN_SET(project,
'hi.wikipedia,bn.wikipedia,id.wikipedia,ar.wikipedia,mr.wikipedia,fa.wikipedia,sw.wikipedia,tl.wikipedia,zh.wikiquote,th.wikipedia,arz.wikipedia,ml.wikipedia,ta.wikipedia,kn.wikipedia,pt.wiktionary,az.wikipedia,gu.wikipedia,ky.wikipedia,sq.wikipedia,ms.wikipedia'
) > 0, view_count, 0)) AS mh_views
  FROM wmf.virtualpageview_hourly 
  WHERE (year = 2018 AND month >=4) OR (year = 2019)
  GROUP BY year, month, day
  ORDER BY year, month, day LIMIT 10000;

"
results <- collect(sql(query))
save(results, file="Data/mobile_heavy_previews.tsv")
```

```{r}
mobile_heavy_previews <- read.delim("data/mobile_heavy_previews.tsv", sep = "\t", stringsAsFactors =FALSE)
mobile_heavy_previews$date <- as.Date(mobile_heavy_previews$date, format = "%Y-%m-%d")
```  

Calculate monthly pageviews and YoY changes on mobile-heavy wikis

```{r}
 mh_previews_monthly <- mobile_heavy_previews %>%
    mutate(date = floor_date(date, "month")) %>%
    filter(date <= '2019-04-01') %>% #filter out months with incomplete datea
    select(-c(1,2,3)) %>% 
    group_by(date) %>% 
    summarise(monthly_views = sum(as.numeric(mh_views))/1E6) %>%
    arrange(date) %>%
    mutate(yearOverYear= monthly_views/lag(monthly_views,12) -1)  %>%
    mutate(type = "previews")

tail(mh_previews_monthly)
```

 
## Interactions on Mobile-Heavy Wikis (Pageviews + Seen Previews)

Create chart of interactions broken down by pageviews and previews

```{r}
 mh_interactions <- rbind(mh_pageviews_monthly, mh_previews_monthly) 
```

Create chart of page interactions on mobile heavy wikis

```{r}
  p <- ggplot(mh_interactions, aes(x=date, y = monthly_views, fill = forcats::fct_rev(type))) +  
    geom_col() +
    scale_y_continuous("seen previews (shown for at least 1 sec)\nand pageviews (in millions)") +
    scale_x_date("Date", labels = date_format("%Y-%m"), date_breaks = "3 months") +
    geom_vline(xintercept = as.numeric(as.Date("2018-04-01")),
               linetype = "dashed", color = "blue") +
    geom_text(aes(x=as.Date('2018-04-01'), y=400, label="Page Previews Deployment Completed"), size=4, vjust = -1.2, angle = 90, color = "black") +
    labs(title = "Mobile-heavy wiki pageviews and seen previews per calendar month") +
    ggthemes::theme_tufte(base_size = 14, base_family = "Gill Sans") +
    theme(axis.text.x=element_text(angle = 45, hjust = 1),
          panel.grid = element_line("gray70"),
          legend.position="bottom",
          legend.title=element_blank(),
          legend.text=element_text(size=14))
  
  ggsave(filename="Mobile heavy wiki interactions_StackedBar.png", plot = p, path = 'figures', units = "in", dpi = 192, height = 6, width = 10, limitsize = FALSE)  
p
```

Calculate monthly interactions (sum of pageviews and previews) on desktop

```{r}
  mh_interactions_total <- mh_interactions %>%
    filter(date >= '2018-04-01') %>%#page previews rolled out in April 2018
    group_by(date) %>%
    summarise(interactions = sum(monthly_views)) %>%
    arrange(date) %>%
    mutate(yearOverYear= interactions/lag(interactions,12) -1) 

knitr::kable(mh_interactions_total)
```   
  
